---
title: "NOAA Storm Database, some answers, part of Coursera Data Science Course Reproducbile Research"
author: "Arnout van der Meijden"
date: "5/25/2020"
output: html_document
---

This report was written as part of the Coursera Data Scientist program, course reproducible research, peer assignment number 2.

This report will answer the following two questions:
+ Q1: Across the United States, which types of events (in EVTYPE variable) are most harmful with respect to population health?
+ Q2: Across the United States, which types of events have the greatest economic consequences?
and most importantly show you every step so this research is reproducible. 

# Synopsis
The NOAA storm database possesses a lot of data. 
This document answers the questions which Events are the most harmful to population health and which Events causes the greatest economic consequences. 
The data analysis which is performed can be easily reproduced beccause all the code and all the steps are provided in this document. 
The biggest problem was to match the Events in the records of the NOAA database with the original 48 Events. 
This matching has been automated but is not perfect. 
The conclusion is that only a few Events causes the most damage on health and on damage. 
Most damage on health is done by Excessive Heat and Tornados. 
For property and crop damage Thunderstorm Wind does the most damage. 

# Data Processing
First step is to read the data from the NOAA storm database.
If the file is not yet locally on your harddisk, you can download it at: [Link to download of NOAA storm datbase](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

```{r}
library(data.table)
file <- "repdata_data_StormData.csv.bz2" ## This assumes you got this file in your working directory
Alldata <- fread(file) 
```

Next step is to get only the data we want to answer these two questions. 
Converting to date 
```{r}
Alldata$BGN_DATE <- as.Date(Alldata$BGN_DATE, "%m/%d/%Y %H:%M:%S")
```

ONly use the data after 1996 because only starting in 1996 they have start registering all the eventtypes
```{r}
Dataafter1996 <- Alldata[BGN_DATE > '1996-01-01']
```

## To answer the first question: "Across the United States, which types of events (in EVTYPE variable) are most harmful with respect to population health?"

We will only keep the interesting columns: the EVTYPE, FATALITIES and INJURIES because these two are most harmful to population health. The Date and Remarks are not really necessary for answering the first questions and could be left out but might help us answer other questions later. 
```{r}
library(dplyr)
Stormdataonhealth <-
    Dataafter1996 %>%
      select(BGN_DATE, EVTYPE, FATALITIES, INJURIES, REMARKS)
```

Let's first focus on the events which have caused one or more fatalities. Because these are the most harmful for peoples health.
```{r}
WhichEVdoesmostdamage <-
  Stormdataonhealth %>%
    group_by(EVTYPE) %>%
      summarize(NrFatalities = sum(FATALITIES))

OnlyEVwithFatdamage <-
  WhichEVdoesmostdamage %>%
    filter(NrFatalities > 0)
```  


Let's see how many unique Event Types we have at this moment.
```{r}
UniqueFATEvents <- unique(OnlyEVwithFatdamage$EVTYPE) 
```
We have 109 unique event types

We need to match these Event Types with the orginal set of Event Types

Getting the official orginal list of EVTYPES
List of the orgnial Event Types is on page 6 in this PDF [Link to orginal PDF with Event Types](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

Copy the list from this PDF in Notepad, save this new file as text (EVTypes.txt), put this file in the workdirectory and then loaded this file in R
```{r}
stringEvents <- read.table("EVTypes.txt", sep="\t")
```

So we now have a list of 48 Event Types we want to match with the events of the uniqueFATEvents
```{r}
vectorEvents <- as.vector(unlist(stringEvents['V1']))
vectorEVFAT <- as.vector(unlist(OnlyEVwithFatdamage['EVTYPE']))
vectorEVFAT2 <- lapply(vectorEVFAT, tolower)
```

To do an automated match we use this function and match them
```{r}

library(stringdist)

FindMatch = function(string, stringVector) {
  amatch(string, stringVector, method='jw', maxDist = 15, p=0.1)
}

matchesfat <- lapply((vectorEVFAT2), FindMatch, stringVector=vectorEvents)

matchesfat2 <- as.vector(unlist(matchesfat))

matchesfat3 <- vectorEvents[(matchesfat2)]

matchesfat4 <- data.frame(matrix(unlist(matchesfat3), nrow=109, byrow=T),stringsAsFactors=FALSE)

matchefat4withID <-
  matchesfat4  %>%
    mutate(id = row_number())

## Add ID's to the dataframe and list so I can merge this by ID
OnlyEVwithFatdamagewithID <-
  OnlyEVwithFatdamage %>%
    mutate(id = row_number())

OnlyEVwithFATdamagecompare <- full_join(OnlyEVwithFatdamagewithID, matchefat4withID)
```

```{r}
colnames(OnlyEVwithFATdamagecompare)[4] <- "Event"
```

Let's see how this matching is done for the events with more than 60 fatalities

```{r}
OnlyEVwithFATdamagecompare2 <-
  OnlyEVwithFATdamagecompare %>%
     filter(NrFatalities > 60) %>%
        arrange(desc(NrFatalities))
```

Table which show if this automated matching worked
```{r}
library(xtable)
library(htmltools)
library(htmlTable)
xtfat <- xtable(OnlyEVwithFATdamagecompare2)
htmlTable(xtfat, caption= "Are the EV Types well matched?")
```

Matching went well, except for the TSTM WIND (id = 95) which should be Thunderstrom Wind and not Strong Wind, let's fix that

```{r}
OnlyEVwithFATdamagecompareimproved <- 
  OnlyEVwithFATdamagecompare %>%
    mutate(Event = case_when(
      id == 95 ~ (Event = "Thunderstorm Wind"),
        TRUE ~ as.character(Event)))
```


Now let's create a table based on the Orgininal Event Types with the highest number of fatalities in this period and show only those events which caused more than 450 fatalities
```{r}
OnlyEVwithFATdamagecompareimproved2 <-
  OnlyEVwithFATdamagecompareimproved %>%
    group_by(Event) %>%
      summarize(Fatalities= sum(NrFatalities)) %>%
        filter(Fatalities > 450) %>%
          arrange(desc(Fatalities))

xtfat2 <- xtable(OnlyEVwithFATdamagecompareimproved2)
htmlTable(xtfat2, caption= "Which Events had more than 450 fatalities?")
```


Next step is to get this same information for the number of injuries. 

```{r}
WhichEVdoesmostdamage2 <-
  Stormdataonhealth %>%
    group_by(EVTYPE) %>%
      summarize(NrInjuries = sum(INJURIES))

OnlyEVwithInjdamage <-
  WhichEVdoesmostdamage2 %>%
    filter(NrInjuries > 0)

UniqueInjEvents <- unique(OnlyEVwithInjdamage$EVTYPE) 
vectorEVInj <- as.vector(unlist(OnlyEVwithInjdamage['EVTYPE']))

## First lower casing so we can better match this.
vectorEVInj2 <- lapply(vectorEVInj, tolower)

matchesinj <- lapply((vectorEVInj2), FindMatch, stringVector=vectorEvents)

matchesinj2 <- as.vector(unlist(matchesinj))

matchesinj3 <- vectorEvents[(matchesinj2)]

matchesinj4 <- data.frame(matrix(unlist(matchesinj3 ), nrow=106, byrow=T),stringsAsFactors=FALSE)

matchesinj4withID <-
  matchesinj4  %>%
    mutate(id = row_number())

## Add ID's to the dataframe and list so I can merge this by ID
OnlyEVwithInjdamagewithID <-
  OnlyEVwithInjdamage %>%
    mutate(id = row_number())

OnlyEVwithInjdamagecompare <- full_join(OnlyEVwithInjdamagewithID , matchesinj4withID)

colnames(OnlyEVwithInjdamagecompare)[4] <- "Event"

OnlyEVwithInjdamagecompare2 <-
  OnlyEVwithInjdamagecompare %>%
  filter(NrInjuries > 60) %>%
  arrange(desc(NrInjuries))
```

Let's see how this automated matching worked for the injuries
```{r}
library(xtable)
library(htmltools)
library(htmlTable)
xtinj <- xtable(OnlyEVwithInjdamagecompare2)
htmlTable(xtinj, caption= "Are the Event Types for the Injuries well matched?")
```

In this case we need to change some more items. Let's do this.
```{r}
OnlyEVwithInjdamagecompareimproved <- 
  OnlyEVwithInjdamagecompare %>%
    mutate(OriginalEventTypes = case_when(
       id == 89 ~ (Event = "Thunderstorm Wind"),
       id == 92 ~ (Event = "Thunderstorm Wind"),
       id == 26 ~ (Event = "Dense Fog"),
       id == 30 ~ (Event = "Frost/Freeze"),
       TRUE ~ as.character(Event)))
```

Let's now see which Events have the highest number of injuries. 
```{r}
OnlyEVwithInjdamagecompareimproved2 <-
  OnlyEVwithInjdamagecompareimproved  %>%
   group_by(Event) %>%
     summarize(Injuries = sum(NrInjuries)) %>%
       filter(Injuries > 2500) %>%
        arrange(desc(Injuries))

xtinj2 <- xtable(OnlyEVwithInjdamagecompareimproved2)
htmlTable(xtinj2, caption= "Which Events had the highest number of injuries?")
```

The results for both Fatalities and Injuries are shown in a graph in the Results section of this document. 


## To answer the second question: "Across the United States, which types of events have the greatest economic consequences?"

Let's start with getting only the data we need to answer this question. We still use only the data after 1996 because from 1996 they start recording all the Events.
So we need of course the Events (EVTYPE). the damage to the property(PROPDMG, PROPDMGEXP) and the data on damage on the crops (CROPDMG, CROPDMGEXP) 

```{r}
Stormdataoneconomics <-
  Dataafter1996 %>%
    select(BGN_DATE, EVTYPE, PROPDMG, PROPDMGEXP ,CROPDMG, CROPDMGEXP)
```


The (...EXP) is filled with letters to tell the amout of damage, B and b are billion, m or M is million, etc. So we need to transfrom this letters to numbers so we can multiply. The total damage is the value in PROPDMG or CROPDMG mulitplied by the number in this new prop or crop exvalue column. 

```{r}
Valuesstormdataoneconomics <- 
  Stormdataoneconomics %>%
    mutate(propexvalue = case_when(
      PROPDMGEXP == "B" ~ 1000000000,
      PROPDMGEXP == "b" ~ 1000000000,
      PROPDMGEXP == "M" ~ 1000000,
      PROPDMGEXP == "m" ~ 1000000,
      PROPDMGEXP == "K" ~ 1000,
      PROPDMGEXP == "k" ~ 1000))


Valuesstormdataoneconomics2 <- 
  Valuesstormdataoneconomics %>%
      mutate(cropexvalue = case_when(
    CROPDMGEXP == "B" ~ 1000000000,
    CROPDMGEXP == "b" ~ 1000000000,
    CROPDMGEXP == "M" ~ 1000000,
    CROPDMGEXP == "m" ~ 1000000,
    CROPDMGEXP == "K" ~ 1000,
    CROPDMGEXP == "k" ~ 1000))


Valuesstormdataoneconomics3 <- 
    Valuesstormdataoneconomics2 %>%
          mutate(realpropvalue = (PROPDMG * propexvalue))

Valuesstormdataoneconomics4 <- 
  Valuesstormdataoneconomics3 %>%
      mutate(realcropvalue = (CROPDMG * cropexvalue))
```


### Let's first focus on the most Property damage
```{r}
    MostPropDamage <-
      Valuesstormdataoneconomics4 %>%
          group_by(EVTYPE) %>%
            summarize(PropDamage = sum(realpropvalue)) %>% 
                  filter(PropDamage > 0) %>%
                       arrange(desc(PropDamage))
```

Match the events

```{r}
vectorPropdam <- as.vector(unlist(MostPropDamage['EVTYPE']))

vectorPropdam2 <- lapply(vectorPropdam, tolower)

matchesprop <- lapply((vectorPropdam2), FindMatch, stringVector=vectorEvents)

matchesprop2 <- as.vector(unlist(matchesprop))

matchesprop3 <- vectorEvents[(matchesprop2)]

matchesprop4 <- data.frame(matrix(unlist(matchesprop3), nrow=53, byrow=T),stringsAsFactors=FALSE)

matchesprop4withID <-
  matchesprop4  %>%
    mutate(id = row_number())

## Add ID's to the dataframe and list so I can merge this by ID
MostPropDamagewithID <-
  MostPropDamage %>%
    mutate(id = row_number())

OnlyEVwithPropdamagecompare <- full_join(MostPropDamagewithID, matchesprop4withID)

colnames(OnlyEVwithPropdamagecompare)[4] <- "Event"

OnlyEVwithPropdamagecompare2 <-
  OnlyEVwithPropdamagecompare %>%
    filter(PropDamage > 250000) %>%
      arrange(desc(PropDamage))
```

How did the matching go?

```{r}
xtprop <- xtable(OnlyEVwithPropdamagecompare2)
htmlTable(xtprop, caption= "How did the matching go?")
```

We need to fix three items. Because I have no clue which Event from NOAA could be linked to Landslump, Landslump was added as Event.

```{r}
OnlyEVwithPropdamagecompareimproved <- 
  OnlyEVwithPropdamagecompare %>%
  mutate(Event = case_when(
    id == 6 ~ (Event = "Frost/Freeze"),
    id == 14 ~ (Event = "Heavy Snow"),
    id == 15 ~ (Event = "Landslump"),
    TRUE ~ as.character(Event)))
```

Let's found out which Evens had the most property damage
```{r}
WhichEventwithmostPropDamage <- 
  OnlyEVwithPropdamagecompare %>%
    group_by(Event) %>%
      summarize(PropertyDamage = sum(PropDamage)) %>% 
         filter(PropertyDamage > 250000) %>%
            arrange(desc(PropertyDamage))


xtprop2 <- xtable(WhichEventwithmostPropDamage)
htmlTable(xtprop2, caption= "Which Event did the most damage to properties?")
```


### Now we know, let's see which Event did the most damage to the crops.

```{r}
MostCropDamage <-
  Valuesstormdataoneconomics4 %>%
    group_by(EVTYPE) %>%
      summarize( CropDamage = sum(realcropvalue)) %>% 
        filter(CropDamage > 0) %>%
          arrange(desc(CropDamage))
```

Only 10 records for Crop Damage. We match this manualy.

```{r}
MostCropDamagewithEvent <- 
  MostCropDamage %>%
    mutate(Event=EVTYPE)

MostCropDamagewImproved <- 
  MostCropDamagewithEvent %>%
  mutate(Event = case_when(
    EVTYPE == "THUNDERSTORM WIND" ~ (Event = "Thunderstorm Wind"),
    EVTYPE == "Early Frost"  ~ (Event = "Frost/Freeze"),
    EVTYPE == "Damaging Freeze"  ~ (Event = "Frost/Freeze"),
    EVTYPE == "AGRICULTURAL FREEZE" ~ (Event = "Frost/Freeze"),
    EVTYPE == "Freeze"  ~ (Event = "Frost/Freeze"),
    EVTYPE == "UNSEASONAL RAIN" ~ (Event = "UNSEASONAL RAIN"),
    EVTYPE == "Unseasonable Cold" ~ (Event = "Frost/Freeze"),
    EVTYPE == "Heavy Rain/High Surf"  ~ (Event = "High Surf"),
    EVTYPE == "Frost/Freeze" ~ (Event = "Frost/Freeze"),
    EVTYPE == "MARINE THUNDERSTORM WIND"  ~ (Event = "Marine Thunderstorm Wind"),
    TRUE ~ as.character(Event)))
```

Let's see which event did the most damage to the crops.

```{r}
WhichEventwithmostCropDamage <- 
  MostCropDamagewImproved %>%
  group_by(Event) %>%
  summarize(CropDamage = sum(CropDamage)) %>% 
      arrange(desc(CropDamage))


xtCrop <- xtable(WhichEventwithmostCropDamage)
htmlTable(xtCrop, caption= "Which Event did the most damage to Crop?")
```


# Results

### Q1: Across the United States, which types of events (in EVTYPE variable) are most harmful with respect to population health?

To make a more visual presentation of the data, see below a graphic which shows what Events are the most harmful. First it will show the Events which caused the most deaths, below that one it will show the events with the most number of injuries. 
```{r plot1}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(data=OnlyEVwithFATdamagecompareimproved2, aes(Event, Fatalities, fill = Event)) + geom_bar(stat = "identity")
p2 <- ggplot(data=OnlyEVwithInjdamagecompareimproved2, aes(Event, Injuries, fill = Event)) + geom_bar(stat = "identity")
grid.arrange(p1, p2, nrow = 2, top='Which Events (based on NOAA) are the most harmful with respect to population health')
```

#### Conlusion: The most harmful Events for the population health are Excessive Heat and Tornado. 

### Q2: Across the United States, which types of events have the greatest economic consequences?
Answer, see the plots below
```{r plot2}
library(treemap)
treemap(WhichEventwithmostPropDamage,
        index=c("Event"),
        vSize = "PropertyDamage",
        type = "index",
        palette = "Reds",
        title="Total Property Damage by Event Type",
        fontsize.title = 14)
```


```{r plot 3}
library(treemap)
treemap(WhichEventwithmostCropDamage,
        index=c("Event"),
        vSize = "CropDamage",
        type = "index",
        palette = "Blues",
        title="Total Property Damage by Event Type",
        fontsize.title = 14)
```

#### Conclusion: Thunderstorm wind is the Event with the greatest economic consequences
