---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This report will answer the questions which are part of this assignment


## Loading and preprocessing the data

To load the data we must first unzip the data and create the data directory
````{r}
unzip("activity.zip", exdir = "data")
```

Next we can read the data

```{r} 
directory <- getwd()
data <- read.csv(paste0(directory,"/data/activity.csv"), sep=",")
```

## What is mean total number of steps taken per day?
To create the histogram of the total number of steps taken each day we need to get this data

```{r message=FALSE} 
library(dplyr)
```

```{r} 
Sumperday <-
  data %>%
    group_by(date) %>%
      summarize(Sumeveryday = sum(steps, na.rm = TRUE))
```

The histogram of the total number of steps taken each day

```{r histogram, fig.height=4, echo=FALSE} 
library(ggplot2)
ggplot(data = Sumperday, aes(Sumeveryday)) + geom_histogram(binwidth = 245, col="blue") + theme_light(base_size = 12) + scale_x_continuous(name = "Total number of steps taken each day")
```


## What is mean total number of steps taken per day?

First we need to calculate this data
```{r} 
MeanMedianperday <-
      data %>%
          group_by(date) %>%
             summarize(mean = mean(steps), median = median(steps)) 
```

To create a table which is well readable we need to load some libraries 
```{r message=FALSE} 
library(xtable)
library(htmltools)
library(htmlTable)
```

To make the table even better we only want to show the number with only 2 digits
```{r} 
RoundedMeanMedianperday <-
  MeanMedianperday %>%
  mutate(meanround = round(mean, digits = 2))

DatafortableRoundedmeanmedianperday <- select(RoundedMeanMedianperday, -mean)
```

The table with median and mean total number of steps taken per day
```{r} 
xt <- xtable(DatafortableRoundedmeanmedianperday)
htmlTable(xt, ctable=c("date", "meanround", "median"), caption= "Table with mean and median numbers of steps each day")
```


## What is the average daily activity pattern?

To make the plot we first need to convert the interval to minutes

```{r} 
tempa <- c(data$interval)
tempb <- sprintf("%04d", tempa)
data2 <- mutate(data, time = (format(strptime(tempb, format="%H%M"), format = "%H:%M")) )
```


Calculate the average number of steps for every 5 minutes
```{r} 
dataonlyintervalavg <-
    data2 %>%
      group_by(time) %>%
        summarize(avgsteps = mean(steps, na.rm = TRUE))
```
 

```{r} 
dataonlyintervalavg$time = as.POSIXct(hms::parse_hm(dataonlyintervalavg$time))
```

The  time series plot (type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r plot, fig.height=4, fig.width=4, echo=FALSE} 
plot(dataonlyintervalavg$time, dataonlyintervalavg$avgsteps, type = "l", col="blue", xlab = "time", ylab = "Average steps")
```


```{r echo=FALSE} 
dataonlyintervalavg <-
    data2 %>%
      group_by(time) %>%
        summarize(avgsteps = mean(steps, na.rm = TRUE))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r} 
maxavgstepstime <- dataonlyintervalavg[which.max(dataonlyintervalavg$avgsteps), ] 
print(maxavgstepstime[1])
```


## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs 
```{r} 
NAinthisset <-is.na(data)
table(NAinthisset)
```

Only for the steps column
```{r} 
NAinsteps <- is.na(data$steps)
table(NAinsteps)
```
So we only miss values in the steps column

The strategy I have choosen for filling in all of the missing values in the dataset is to fill with the value of the mean of that 5-minute interval

To replace the NA's with the mean of the 5-minute interval
Add interval to the dataonlyyintervalavg dataset (this dataset has the average)

```{r} 
temp <- c(data$interval)
interval1day <- temp[1:288]

  doiavg2 <-
      dataonlyintervalavg %>%
        mutate(interval = interval1day)
  
```

Merging the intitial dataset with this new dataset to get for every interval also the 5 minute mean
```{r} 
newdata3 <- inner_join(data, doiavg2, by=c("interval"))
```
Imputing the missing data with the average of that time
```{r} 
newdata4 <-
  newdata3  %>%
    mutate(steps=ifelse(is.na(steps),(avgsteps), steps))
```
Removing all unnecessary columns
```{r} 
newdata5 <- subset(newdata4, select = c(steps, date, interval))
```

Are there any NAs in this new dataset
```{r} 
NAinnewdataset5 <-is.na(newdata5)
table(NAinnewdataset5)
```

Histogram of the total number of steps taken each day
```{r histogram2, fig.height=4, echo=FALSE} 
library(ggplot2)
ggplot(data = newdata5, aes(steps)) + geom_histogram(col="green") + scale_fill_gradient("count", low="green", high="red")
```

Calculate and report the mean and median total number of steps taken per day. 
Also we round this time both the mean and median to make the table look better
```{r} 
MeanMedianperdaynewdata   <-
    newdata5 %>%
      group_by(date) %>%
          summarize(mean = mean(steps), median = median(steps))

RoundedMeanMedianperdaynewdata <-
 MeanMedianperdaynewdata %>%
  mutate(meanround = round(mean, digits = 2)) %>%
    mutate(medianround = round(median, digits = 2))

DatafortableRoundedmeanmedianperdaynewdata <- select(RoundedMeanMedianperdaynewdata, -c(mean, median))

xt2 <- xtable(DatafortableRoundedmeanmedianperdaynewdata)
htmlTable(xt2, ctable=c("date", "meanround", "medianround"), caption= "Table with mean and median numbers of steps each day after imputting missing values")
```

If we compare the new values with the old then it is clear that we don't have any NA values anymore but the average and mean of a specific day is the same. Except of course if the previous data was NA.
Another observation is that only for the dates we filled with data, there is now also a Median > 0. Of course this values is the same for all these dates because the missing data was for complete dates. 

## Are there differences in activity patterns between weekdays and weekends?

Converting to a date format
```{r echo=FALSE} 
library(lubridate)     
newdata5$date <- ymd(newdata5$date)
```

Getting the days of the week using the weekdays 
```{r} 
newdata6 <-
    newdata5  %>%
       mutate(weekday = weekdays(newdata5$date))
```

Converting the days to weekend or weekday
```{r} 
newdata7 <-
    newdata6 %>% 
      mutate(weekendornot = ifelse(newdata6$weekday %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


Add the time to the dataset 
```{r echo=FALSE} 
tempa <- c(data$interval)
tempb <- sprintf("%04d", tempa)
newdata71 <- mutate(newdata7, time = (format(strptime(tempb, format="%H%M"), format = "%H:%M")))
```

Then split the data in weekday and weekend
```{r} 
newdata8 <- split(newdata71, newdata71$weekendornot)
```


The average number of steps for every 5 minutes for weekday
```{r} 
dataonlyintervalavgweekday <-
  newdata8[[1]] %>%
    group_by(time) %>%
        summarize(avgsteps = mean(steps, na.rm = TRUE))
```

The average number of steps for every 5 minutes for a weekend
```{r} 
dataonlyintervalavgweekend <-
  newdata8[[2]] %>%
      group_by(time) %>%
          summarize(avgsteps = mean(steps, na.rm = TRUE))
```

Make time of time for the plot
```{r} 
dataonlyintervalavgweekday$time = as.POSIXct(hms::parse_hm(dataonlyintervalavgweekday$time))
dataonlyintervalavgweekend$time = as.POSIXct(hms::parse_hm(dataonlyintervalavgweekend$time))
```

The plot
```{r plot3, fig.height=10, echo=FALSE}  
par(mfrow=c(2,1))
plot(dataonlyintervalavgweekend$time, dataonlyintervalavgweekend$avgsteps, type = "l", col="blue", xlab = "time", ylab = "Average steps in the weekend", main ="weekend")
plot(dataonlyintervalavgweekday$time, dataonlyintervalavgweekday$avgsteps, type = "l", col="blue", xlab = "time", ylab = "Average steps on a weekday", main = "weekday")
```

As the graphs show there is a signifcant difference in activity during a weekday or in the weekend, as we would expect.
